---
title: 'Exam'
author: 'Vivek Panchal'
date: '2019-10-13'
output:
    pdf_document:
    latex_engine: xelatex
header-includes:
  \usepackage{booktabs}
---

```{r echo=FALSE}
# This chunk is just to make it possible to shrink the typeface in succeeding chunks. Mainly this will be used for the crosstabs.
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

# 1. Introduction
H1-B is an employement based visa category for non-immigrants who are temporary foreign workers in United States.The original dataset is obtained from U.S Department of Labor. The dataset used for analysis is fitered from October 2016 till June 2017.
  
## 1.1 Loading and installing required libraries
```{r Installing required packages}
library(ggplot2)
library(tidyverse)
library(tidyr)
install.packages('tidyverse', repos = "http://cran.us.r-project.org")
library(data.table)
library(ggmap)
library(knitr)
library(ggplot2)
library(dplyr)
```

## 1.2 Reading csv

```{r Initialize}
h1b<-fread("C:/Users/vivek 14/Downloads/h1bdata.csv")
```

## 2. Cleaning H1BDataset
In order to clean the dataset, first, we take the loaded dataset and omit "NA" values from it.

```{r Data Cleaning}
h1b<-na.omit(h1b)
h1b<-distinct(h1b)
```

## 3. Data Exploration
In this dataset, we have a column name **Case_Status** which indicates the status associated with the petition filed. Below a representation of different statuses is shown for the year 2016 and 2017.
```{r Representation of various statuses}

h1b_plot1<-ggplot(subset(h1b, !is.na(h1b$CASE_STATUS)), 
       aes(x = CASE_STATUS, fill = CASE_STATUS)) +
    geom_bar(stat = "count") +
    coord_trans(y = "sqrt") +
    ylab("Applications Filed") +
    theme(legend.position = c(0.9, 0.9), 
         legend.key.size = unit(0.3, "cm"),
          axis.text.x=element_text(angle = -10, hjust = 0, size = rel(1))) +
    ggtitle("H1B visa case status")
h1b_plot1
```

# 3.1 Filtering values for analysis
Here we filter the dataset with respect to status as **Certified**. This filtration helps in the analysis which is shown below.

```{r filtering}
certified_h1b <- h1b %>%
    filter(CASE_STATUS == "CERTIFIED")
state_h1b<- h1b[,c(25)]
```

# 3.2 Popular H1B Visa Sponsors
To dive deeper into the analysis, it is very important to know which companies file the most number of H1B petitions. The visualization below indicates such results.

```{r top employers}
topemployers <- function(num_emp) {
    certified_h1b %>%
        group_by(EMPLOYER_NAME) %>%
        summarise(num_apps = n()) %>%
        arrange(desc(num_apps)) %>%
        slice(1:num_emp)
}
h1btopemployers<-ggplot(topemployers(10), 
       aes(x = reorder(EMPLOYER_NAME, num_apps), y = num_apps)) +
    geom_bar(stat = "identity", alpha = 0.9, fill = "#228B22", width = 0.7) +
    coord_flip() +
    scale_y_continuous(limits = c(0, 20000), breaks = seq(0, 20000, 2000)) +
    geom_text(aes(label = num_apps), hjust = -0.2, size = 2) +
    ggtitle("Top 10 Employers with most applications") +
    theme(plot.title = element_text(size = rel(1)),
          axis.text.y = element_text(size = rel(1.1))) +
    labs(x = "Employer Name", y = "No. of Applications") 
h1btopemployers
```

# 3.3 Statewise Comparision
There are states in the United States where people love to work. So from our dataset, we pick three states such as California, New York and District of Columbia and obtain a pictorial representation of the comparison being made. We also get to know the vast differences in the petitions being filed for three states.

```{r Cases in three states }
cnt_case_per_year <- h1b %>%
    filter(EMPLOYER_STATE %in% c("NY", "CA", "DC")) %>%
    group_by(DECISION_YEAR,EMPLOYER_STATE) %>%
    summarise(count = n()) %>%
    arrange(DECISION_YEAR,EMPLOYER_STATE) 


statewise<-ggplot(cnt_case_per_year, aes(x = DECISION_YEAR, y = count, fill = EMPLOYER_STATE)) +
    geom_bar(stat = "identity", position = position_dodge(), alpha = 0.8,
             color = "grey") +
    ggtitle("H1B cases in California, New York and District of Columbia") +
    theme(legend.position = "bottom",
         plot.title = element_text(size = rel(1.4)))
statewise
```

# 3.4 Total Counts
Total count of H1B petitions filed in different states of the USA and top 10 job roles which had the most number of cases filed.

```{r Count of top employers and top job roles }
top<-h1b %>% dplyr::group_by(EMPLOYER_STATE) %>% dplyr::summarise(n = n())
ordered_top <- top[order(-top$n),]
kable(ordered_top)

top_jobs<-h1b %>% count(SOC_NAME)
ordered_top_jobs <- top_jobs[order(-top_jobs$n),]
ordered_top_jobs<-head(ordered_top_jobs,10)
kable(ordered_top_jobs)
```

In the exploration below we obtain the wages of applicants with case status being either denied or certified or withdrawn. The prevailing wages of denied H1B cases have more extreme values than those of certified H1B cases. 

```{r wages of employee with status type}
h1b_status <- h1b %>%
    filter(CASE_STATUS == "CERTIFIED"|CASE_STATUS == "DENIED"|CASE_STATUS == "WITHDRAWN")


h1b_c_d<-ggplot(h1b_status, aes(y = PREVAILING_WAGE, x = CASE_STATUS, 
                                 fill = CASE_STATUS, notch = TRUE, 
                                 notchwidth = .3)) + 
      geom_boxplot(notch = TRUE) + 
      scale_fill_manual(values = c("#29a329", "#ff0831","#3349FF"))+
      scale_y_continuous(limits = c(0, 150000), 
                         breaks = seq(0, 150000, 10000)) + 
      ggtitle("Wages for certified,denied and withdrawn H1B cases")+
      theme(
            plot.title = element_text(size = rel(1.3)),
            panel.background = element_rect(fill = 'white'),
            panel.grid.major = element_line(colour = '#f0f0f0'),
            panel.grid.major.x = element_line(linetype = 'blank'),
            panel.grid.minor = element_line(linetype = 'blank')
      )
h1b_c_d
```

Here we get the top 15 occupations with the highest median prevailing wages. Using the median wage as a metric helps in avoiding distortions and can provide better visualization.

```{r top 15 jobs and their median salary}
top_15_soc_highest_wage <- h1b %>%
  group_by(SOC_NAME) %>%
  summarise(median_wage = median(PREVAILING_WAGE)) %>%
  arrange(desc(median_wage)) %>%
  slice(1:15) %>%
  select(SOC_NAME, median_wage)

ggplot(top_15_soc_highest_wage,aes(x = SOC_NAME, y = median_wage)) +
  geom_bar(stat = "identity", fill = "#40e0d0", alpha = 0.9,  width = 0.7) +
  coord_flip() +
  ggtitle("Top 15 Occupation and their median pay") +
  labs(x = "Occupations", y = "median salary")
```

Here we get the count for the number of an application filed for the corresponding year with an addition to the full-time position obtained or not.

```{r Count of applicants }
h1b %>%
  group_by(FULL_TIME_POSITION, DECISION_YEAR) %>%
  summarise(Num_applications = n())
```

## 4. Segregation of Job types.

Next, a stacked bar chart to investigate the trend of the proportion of H1B applicants who have computer occupations from 2016 to 2017 and compared with other occupation. The result obtained for computer occupations over the total occupations gradually increased over the years. This indicates the flourish in the tech world and interests of people more aligned to computer-related jobs.

```{r Comparision of job types}
case_quantity_per_year <- certified_h1b %>%
    group_by(DECISION_YEAR) %>%
    summarise(all_occupations = n())
case_quantity_per_year

cmo_quantity_per_year <- certified_h1b %>%  filter(SOC_NAME =='COMPUTER OCCUPATION'& !is.na(PREVAILING_WAGE)) %>%
    group_by(`DECISION_YEAR`) %>%
    summarise(Comp_occupation = n(),
           Comp_wages = median(PREVAILING_WAGE))
cmo_quantity_per_year

multi_df1 <- merge(x = cmo_quantity_per_year,
                   y = case_quantity_per_year,
                   by = "DECISION_YEAR",
                   all = TRUE)


multi_df1 <- multi_df1 %>%
    mutate(other_occupations = all_occupations - Comp_occupation,
           Comp_percent = Comp_occupation / all_occupations) %>%
    select(DECISION_YEAR, Comp_occupation, other_occupations)

multi_df1 <- gather(multi_df1, occupation, count, Comp_occupation:other_occupations)

ggplot(multi_df1,
       aes(x = DECISION_YEAR, y = count, fill = occupation)) +
    geom_bar(stat = "identity") +
    scale_y_continuous(breaks = seq(0, 600000, 100000), 
                       limits = c(0, 600000), labels = scales::comma) +
    labs(x = "Year", y = "No. of H1B Cases", 
         title = "Change in the proportion of computer and other occupations")
ggplot

```

# 4.1 Comparision in wages.

There is an obvious increase in the wages for computer-related occupations. There are some fluctuations for other occupations having no significant increase.

```{r Comparision of job wages}
multi_df2 <- certified_h1b %>%
    filter(!is.na(PREVAILING_WAGE)) %>%
       mutate(occupation_category = if_else(SOC_NAME =='COMPUTER OCCUPATION'|SOC_NAME =='ANALYSTS'|SOC_NAME =='ACCOUNTANTS', 
                                         "Comp_occupation", "other_occupations")) %>%
    select(DECISION_YEAR,occupation_category,PREVAILING_WAGE)


colors = c(rep("red"),rep("blue"))
cmo_boxplot <- ggplot(multi_df2, aes(x=occupation_category, y= PREVAILING_WAGE)) +
    geom_boxplot(col=colors,aes(fill=DECISION_YEAR)) + xlab("Occupation Category") + ylab("Wage (USD)") +
    ggtitle("Wages of Computer Occupations and Other Occupations") +
    coord_cartesian(ylim=c(25000,150000))

cmo_boxplot

```

## 5. Conclusion
The above analysis helps in answering questions with respect to H1B cases being filed in the year 2016 and 2017. The first analysis shows us different types of statuses represented in different colours with respect to applications filed. The second analysis tells us which companies file the most number of applications for the readability purpose top 15 companies were picked. In the succeeding analysis we found out the significant increase in the number of applications filed for the California state. This tells us that it is one of the favourite places where people love to work. After that, we get to know the count of top 10 job categories for which applications were filed and count of applications in each state. Later, comparisons were made on the basis of occupations and difference in wages of computer-related occupations and other occupations. This portrays that the tech industry is evolving with every passing year.The number of people wanting to work in fields related to computer as compared to others. There is also a significant increase in pay scale working in the tech industry with comparison to other occupations.






