---
title: 'homework vi'
author: 'Vivek Panchal,Rohit Kunjilikattil'
date: '2019-10-11'
output:
    pdf_document:
    latex_engine: xelatex
header-includes:
  \usepackage{booktabs}
---

```{r echo=FALSE}
# This chunk is just to make it possible to shrink the typeface in succeeding chunks. Mainly this will be used for the crosstabs.
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

# 1. Introduction

This an an exploratory analysis of the NYC311 dataset. Firstly, the dataset is cleaned so that it is ready for use. It includes various visualization on borough against complaint types and against agencies. Another dataset, the NYC_Payroll dataset is cleaned and introduced. A join is performed on the summarized version of NYC311 and the NYC_Payroll database. Lastly, it includes visualization in line with the joins performed.  

## 1.1 Loading and installing required libraries
```{r INSTALLING REQUIRED PACKAGES}
library(ggplot2)
library(dplyr)
library(tidyverse)
library(tidyr)
install.packages('tidyverse', repos = "http://cran.us.r-project.org")
library(data.table)
library(ggmap)
library(knitr)
library(lubridate)
library(magrittr)
library(data.table)
library(gender)
library(dplyr)
library(ggjoy)
library(ggridges)
library(gridExtra)
library(ggthemes)
library(reshape2)
library(corrplot)
```

Tidyverse is a r package that helps in making data tidy. Tidy data is defined as data which follows three basic rules : each variable must have its own column, each observation must have its own row and finally each value must have its own cell.One of the advantages of tidy data is it's easier to work with data that always has a consistent structure. Further, R's vectorised nature makes it particularly easy to transform tidy data.

## 1.2 Reading csv

```{r Initialize}
nyc311<-fread("C:/Users/vivek 14/Downloads/311_Service_Requests_from_2010_to_Present.csv", nrows = 100000)
options("scipen" = 100,"digits" = 4)
```

# 2. Cleaning NYC311Data

Cleaning NYC311 data to save  processing power on bad data. As the complaint type is one of the most important columns,it is made as uniform as possible. We go about doing this byOther cleaning include removing any unwanted spaces, replacing hyphens with slashes and finally filtering the zipcodes to only have values of length 5. 

```{r Data Cleaning}
nyc311$`Complaint Type` = tolower(nyc311$`Complaint Type`) 
nyc311$`Complaint Type` = gsub('s$', '', nyc311$`Complaint Type`) 
nyc311$`Incident Zip` = gsub('-[[:digit:]]{4}$', '',nyc311$`Incident Zip`)
nyc311$`Complaint Type` = gsub('paint - plaster', 'paint/plaster', nyc311$`Complaint Type`)
nyc311$`Complaint Type` = gsub('general construction', 'construction', nyc311$`Complaint Type`)
nyc311$`Complaint Type` = gsub('nonconst', 'construction', nyc311$`Complaint Type`)
nyc311$`Complaint Type` = gsub('street sign - [[:alpha:]]+', 'street sign', nyc311$`Complaint Type`)
nyc311$`Complaint Type` = gsub('fire alarm - .+','fire alarm', nyc311$`Complaint Type`)
nyc311_clean2 = nyc311
```

## 2.1 Separating Columns

The data contains multiple columns that are of class character that have date and time values together in one column. The **separate** function from tidyr library helps in separating such columns into columns with date and time that are stored separately.

```{r Creating seperate Columns for better analysis of data}
nyc311_clean2 <- nyc311_clean2 %>% 
  separate(`Created Date`, into = c("Created Date", "Created Time"), sep = 10)

nyc311_clean2 <- nyc311_clean2 %>% 
  separate(`Closed Date`, into = c("Closed Date", "Closed Time"), sep = 10)

nyc311_clean2 <- nyc311_clean2 %>% 
  separate(`Due Date`, into = c("Due Date", "Due Time"), sep = 10)

nyc311_clean2 <- nyc311_clean2 %>% 
  separate(`Resolution Action Updated Date`, into = c("Resolution Action Updated Date", "Resolution Action Updated Time"), sep = 10)
```

# 3. Data Exploration of NYC311 Dataset

## 3.1 Exploration using Latitude and Longitude columns in data.

Location data is very helpful in visualization. The ggmap library in order **to find out which locations have complaints with status as "CLOSED".** The filtering is done so that the exploration gives us a sense of efficiency of the oraganization in different boroughs

Above Map is made with data that contains the selected four columns `Borough`,`Latitude`,`Longitude`,`Status`. The map used is "terrain".Some data contains rows with which have 'Boroughs' as 'Unspecified'. These rows are removed. For better readability, different boroughs are marked with different colors.

In order to visualize maps, library "ggmap" is used . With the help of this package, it is easy to retrieve map tiles from sources such as Google Maps, Open Street Maps and many more. They have the necessary tools which helps in smooth functioning for our routing. The two columns which are very essential to visualize any maps are  `Latitude` and `Longitude`. 

For the purpose of this visualization you need to enable google static map service. If this service was to be used before July 2017,  there was no need to signup for any google map services. But now it requires setting up an API Key and enabling billing. While signing up for the service you need to copy your API key which will be added to your chunk.

```{r EXPLORING MAP DATA}

nyc311_map<-nyc311[sample(nrow(nyc311),10000),]
nyc_map_data <- nyc311_map %>% select(24,50,51,20)

nyc_map_data<- nyc_map_data %>% filter(!str_detect(Borough, "Unspecified"))
nyc_map_data<-na.omit(nyc_map_data)
xtab<-dplyr::filter(nyc_map_data,Status == "Closed")

#library(devtools)
#devtools::install_github("fresques/ggmap")
# devtools::install_github("dkahle/ggmap")

key<-'AIzaSyD8Zu8HBooWB2MP9yHqbXOA5ERj5oTQ-mA'
register_google(key ="AIzaSyD8Zu8HBooWB2MP9yHqbXOA5ERj5oTQ-mA")
nyc_map = get_map(location = c(lon = -73.9, lat = 40.7),
                  zoom = 11, maptype="terrain")
map<-ggmap(nyc_map) + geom_point(data=xtab,aes(x=xtab$Longitude,y=xtab$Latitude,shape =factor(Borough),color=factor(Borough)),size=2,alpha=0.2) + ggtitle("MAP") + theme(plot.title=element_text(hjust=0.5)) + xlab('Longitude') + ylab('Latitude')
map

```

## 3.2 Exploring relationship between complaints in different Borough

Next, an exploration to demonstrate the relationship between top complaints and the boroughs. This would give an insight about what complaints are registered more frequently in which boroughs. 

The first figure is barplot of the top complaints to showcase the most registered complaints in the nyc311 data. The second graph is an an overview of distribution of complaints across different boroughs, while the third graph is a closer look at the situation in each borough.The third graph is divided by the color scheme based on agencies. This can help in showing which agency is responsible for handling which type of complaint.


```{r Exploration of data}
all_complaints <-  nyc311 %>% select(2, 6, 24)

all_complaints$`Complaint Type`[grepl("^Noise.*", all_complaints$Complaint.Type)] <-
"Noise"

all_complaints_temp <- all_complaints %>%
group_by(`Complaint Type`) %>%
summarise(count = n()) %>%
arrange(desc(count))
top10_complaints <- top_n(all_complaints_temp, 10, count)
ggplot(top10_complaints) + geom_bar(
aes(x = top10_complaints$`Complaint Type` , y = top10_complaints$count),
fill = "Orange",
stat = "identity"
) + theme_minimal() +
xlab("") + ylab("Number of Complaints") + coord_flip()


nyc311_subset <- subset(nyc311,`Complaint Type` %in% count(nyc311,`Complaint Type`,sort = T)[1:10,]$`Complaint Type` )
nyc311_subset <- nyc311_subset %>% select(`Complaint Type`,Borough,Status,Agency)  %>% group_by(Borough) 
nyc311_subset <- nyc311_subset %>% filter(!str_detect(Borough, "Unspecified"))
ggplot(nyc311_subset, aes(`Complaint Type`)) + geom_bar(stat = "count")  + labs(x = "Top 10 Complaint Types",
                              y = "Number of complaints",
                              title = "Top Complaints in NYC311 Service Requests by Borough",
                              subtitle = "")  +coord_flip()  + aes(fill=Borough)

ggplot(nyc311_subset, aes(`Complaint Type`)) +
  geom_bar(stat = "count") +
  facet_wrap(~Borough) + labs(x = "Top 10 Complaint Types",
                              y = "Number of Complaints",
                              title = "Top Complaints in NYC311 Service Requests by Borough ",
                              subtitle = "") + coord_flip()  + aes(fill=Agency) + theme(axis.text.x = element_text(angle = 90))

```

# 4. Reading a dataset from a url

Next, the NYC_Payroll data is read directly from the url to work on it without having to download the csv file. This dataset contians the salary, pay rate and total compensation of every New York City employees from year 2014 till 2017.This dataset provides columns for fiscal year, agency they work for, borough they are working in. Some useful explorations as who gets paid how much and for what are done. The source dataset thatis being extracted from url follows the rules of tidydata.

```{r Reading csv from URL}
nyc_payroll <- fread("https://data.cityofnewyork.us/api/views/k397-673e/rows.csv?accessType=DOWNLOAD")
nyc_payroll<- nyc_payroll[which (nyc_payroll$`Work Location Borough` == 'MANHATTAN' |nyc_payroll$`Work Location Borough` == 'BRONX'|nyc_payroll$`Work Location Borough` == 'BROOKLYN' |nyc_payroll$`Work Location Borough` == 'QUEENS'|nyc_payroll$`Work Location Borough` == 'STATEN ISLAND')]
nycpayroll_nodups<-distinct(nyc_payroll)
isTRUE(all.equal(nycpayroll_nodups,nyc_payroll))
```

# 5. Exploring NYCPayroll Dataset

There are 3 types of pay basis i.e Per hour, Per day and Per annum. Exploration of it is done by finding the total count of such employees working in New York city.

```{r Payment Categories}
nyc_payroll %>% dplyr::group_by(`Pay Basis`) %>% dplyr::summarise(n = n())

```

The total count of records by year and displaying first five records of the result set.

```{r Count of records by year}
nyc_payroll %>% dplyr::group_by(`Fiscal Year`) %>% dplyr::summarise(n = n())

#Showing first five records of the nyc_payroll dataset.
nyc_payroll %>%
  head(5)%>%
  kable()
```

The dataset is split into two parts, agencies that have income less than 10k and over 10k.For representation of agencies that have **Base Salary** more than 10K use of kable() of knitr library is done to display data in concise manner

```{r Agency representation under 10k}

byagencysub10<- nyc_payroll %>% count(nyc_payroll$`Agency Name`, nyc_payroll$`Fiscal Year`) %>% ungroup() %>% arrange(desc(n))
kable(head(byagencysub10,10))

#Agency representation over 10k

byunion10plus<- nyc_payroll %>% count(nyc_payroll$`Agency Name`,nyc_payroll$`Title Description`,nyc_payroll$`Base Salary`) %>% ungroup() %>% arrange(desc(n))
kable(head(byunion10plus,10))
```

Calculation of the average salary of each agency is done in terms of `Regular Gross Paid`. For this analysis, dplyr functions are used. Group by is done on the `Fiscal year` column of nycpayroll dataset.Mutate() helped in creating a new variable from the data. The summarise_at() affects variables selected with character vector or vars().vars() is nothing but a character vector of column names, a numeric vector of column positions.

```{r Avg salary by Agency}
salbyagency <- nyc_payroll %>%
  group_by(`Fiscal Year`) %>%
  mutate(Count = n()) %>%
  group_by(`Agency Name`, `Count`) %>%
  summarise_at(vars(`Regular Gross Paid`), funs(mean(., na.rm = TRUE))) %>%
  arrange(desc(`Regular Gross Paid`))
kable(head(salbyagency, 10))
```

In our dataset we have a column `Agency Start Date` so what we did is we created a new column `yrs` from it which has the year of that particular person when he started working. We made use of POSIXct for this purpose. Further to calculate experience we took out present date and from that we subtracted agency start date.We performed group by on Boroughs so that we get average years of experience of each Borough.

```{r summarize by borough and calculate average years of experience from Start Date - }

nyc_payroll$`yrs` <- year(as.POSIXct(nyc_payroll$`Agency Start Date`, format="%d/%m/%Y"))
today <- format(Sys.Date(), "%Y")
nyc_payroll$exp <- as.numeric(today)-as.numeric(nyc_payroll$yrs)
salbyexp<- nyc_payroll%>%
  group_by(`Work Location Borough`) %>%
  select(`Work Location Borough`,exp) %>%
  dplyr::summarise_at(vars(`exp`), funs(mean(., na.rm=TRUE)))
kable(head(salbyexp,10))
```

As we are going to join our nyc payroll dataset with nyc311 open data set we have common columns such as Boroughs and Agency for that purpose we renamed our column name of `Work Location Borough` of nyc_payroll dataset to `Borough` so that we can perform merge operation.

```{r Renaming column}
nyc_pr_br<-as.data.table(nyc_payroll) %>% {setnames(., old = "Work Location Borough", new = "Borough")[]}

nyc_payroll %>% filter(`Pay Basis`=='per Annum') %>% filter(`Work Location Borough`!="") %>% group_by(`Work Location Borough`) %>% summarize(count=n()) %>% arrange(-count)
```

# 6. Gender Prediction

R has a library gender which helps in determining gender of a person from the first name. It takes in arguments such as name, years and method. The "ssa" method takes up name from SSN administration of new born baby data.

```{r Gender Classification}

nyc_payroll<-as.data.frame(nyc_payroll)
all_names <- nyc_payroll %>% select(`First Name`) %>% sapply(as.character) %>% as.vector()
min_year<-rep(1950,length(all_names))
max_year<-rep(2012,length(all_names))
temp_df <- data_frame(first_names = all_names,min_year = min_year,max_year = max_year)

results_df <- gender_df(temp_df, name_col = "first_names", year_col = c("min_year", "max_year"), method = "ssa")

#bind input and output
res<-temp_df %>% left_join(results_df, by = c("first_names" = "name", "min_year" = "year_min", "max_year" ="year_max"))

nyc_payroll<-cbind(nyc_payroll, res)
```

## 6.1 Salary Evolution over Years

As the dataset is large,it is hard for data to be distributed normally, thus taking the mean or median of a distribution may lead to different resultsets.

It is observed that over the years, the distribution hasn't changed much. Hence the median value may be more appropriate to describe the salary feature.

```{r}

g1<-nyc_payroll %>% dplyr::filter(`Pay Basis`=='per Annum') %>% na.omit() %>% ggplot(aes(x= `Regular Gross Paid`,y=factor(`Fiscal Year`),fill=gender,height = ..density..)) + geom_joy(scale=1,stat='density',alpha=.7) + theme_joy() + scale_fill_manual(name="",values = c("#3f83a3","#f48000"))+theme(legend.position='top') +theme(axis.text.x =  element_text(angle = 90) )

g2<-nyc_payroll %>% dplyr::filter(`Pay Basis`=='per Annum') %>% na.omit() %>% ggplot(aes(x= `Base Salary`,y=factor(`Fiscal Year`),fill=gender,height = ..density..)) + geom_joy(scale=1,stat='density',alpha=.7) + theme_joy() + scale_fill_manual(name="",values = c("#3f83a3","#f48000")) +theme(legend.position='top') +theme(axis.text.x =  element_text(angle = 90) )

grid.arrange(g1,g2,ncol=2)
```

## 6.2 Breakdown by Borough

Some Boroughs shows data for  only 1 gender, the reason is because of the low statisitic for these Boroughs.
Manhattan being the Borough with most of the data, it proves to be the most trustable.

```{r}
nyc_payroll %>% filter(`Pay Basis`=='per Annum') %>% 
  na.omit() %>%
  group_by(`Fiscal Year`, `gender`, `Work Location Borough`) %>%
  summarize(count = n(), 
            medianSalary = median(`Base Salary`)) %>% 
  ggplot(aes(x=factor(`Fiscal Year`),y=`medianSalary`,group=`gender`)) +
  geom_line(aes(color=`gender`),size=2,alpha=.5) + geom_point(size=1.5) + 
  scale_color_manual(name="",values = c("#3f83a3","#f48000")) + theme_fivethirtyeight() +
  theme(plot.title=element_text(face="bold",hjust=.012,vjust=.8,colour="#3C3C3C",size=14),
        plot.subtitle=element_text(size=10, hjust=0, face="italic", color="black")) + 
        labs(
        title="Median of the Base Salary($US) per Gender over years",
        subtitle="statistics") + facet_wrap(~`Work Location Borough`,ncol = 5)

```

# 7. Joining Dataset

Join is performed between a summarized version of nyc311, summarized on borough and nyc_pr_exp, which was summarized based on average experience in Boroughs. The final result of this join is a table containing the total types of complaint and the total average experience in that borough. This can help in visualizing if the boroughs with a variety of complaints have the required experience to handle it. So more the type of complaints, more should be the average experience.

```{r join}
nyc_pr_exp<-as.data.table(salbyexp) %>% {setnames(., old = "Work Location Borough", new = "Borough")[]}
nyc_ct <- nyc311_clean2[,c(8,28)]
nyc_ct <-  nyc_ct%>% dplyr::group_by(`Complaint Type`) %>% dplyr::summarise(n = n())
nyc_br <- nyc311_clean2 %>% dplyr::group_by(`Borough`) %>% dplyr::summarise(n = n())
merged<- inner_join(all_complaints_temp,all_complaints, by ='Complaint Type')
merged<- merged %>% filter(!str_detect(Borough, "Unspecified"))
    
merged <- merged %>% group_by(Borough, `Complaint Type`) %>% summarise(n = mean(count))
merged <- merged %>% group_by(`Complaint Type`) %>% filter(n()>1)
final_merge <- inner_join(nyc_pr_exp,merged, by='Borough')
final_merge <- final_merge %>% group_by(Borough, `Complaint Type`) %>% summarise(n = mean(exp))
final_merge2 <- final_merge %>% group_by(`Complaint Type`) %>% summarise(n = n())
final_join <- inner_join(final_merge,final_merge2, by ='Complaint Type')

final_join2 <- final_join %>% group_by(`Borough`) %>% summarise(Total_Avg_Exp = mean(n.x), Total_Comp_Types = round(mean(n.y),0))

merged <- distinct(merged,x)
merged <- na.omit(merged)

head(final_join2)
```

This corrplot is the visulaization of the relationship between the Boroughs and the average experience in the boroughs.

```{r Corrplot}
nyc311_corr <- final_join %>% select( 1,3)
new_table2 <- table(melt(nyc311_corr, id.var="Borough"))
new_table2<- as.data.table(new_table2)
nyc311_corr2<-new_table2 %>% select( 1, 4)
wide_table <- spread(new_table2,'Borough',N)
wide_table <- wide_table[,3:6]
resultfinal <- cor(wide_table, use = "complete.obs")
p <- corrplot(resultfinal, type="upper", order="hclust")
p
```

This is a barplot showcasing the different average experience and the total number of complaint types in each borough. It can help to identify the borough which have a variety of complaints and yet have low average experience. The government can take special care to make sure that the complaints in such boroughs are handled efficiently.

```{r Barplot}

test <- rbind(final_join2$Total_Avg_Exp,final_join2$Total_Comp_Types)
bnames <- c("BRONX","BROOKYLN","MANHATTAN","QUEENS")
plot<-barplot(test, beside=TRUE, axisnames=TRUE,
              main = 'Average years of Experience vs Total Complaint Types',
              ylab = 'Count',
              col=c('blue', 'red'),
              names.arg=bnames)

legend("topright", cex=0.6,  c('Avg Exp' , 'Complaint Types'), fill=c('blue', 'red'))

```

# 8. Conclusion

This report is an overview of various exploratory analysis performed on the NYC311 dataset. The analysis shows the differenct relations between aspects of the dataset like, relation between boroughs and complaint types, between complaints and location data and many others. Further, there is an exploration of a relatable dataset, the NYCPayroll dataset. This dataset is summarized into a dataset with borough and the total average experience across the boroughs. It is then joined to a summarized version of the NYC311 dataset containing boroughs and total complaint types. Finally, the report contains some visualizations on the joined data.

# 9. Appendix

1. Fiscal Year - It indicates the tax year in which the job was obtained.

2. Agency Name - The payroll agency the employee works for.

3. Work Location Borough - It indicates employee's work location.

4. Base Salary - Base salary the employee obtains.

5. Regular Gross Paid - Amount paid to employee during fiscal year.

6. Agency Start Date - Date on which employee started working for agency.

7. exp - The average experience of employees of the specified Borough.



